Deploying and Sharing QSAR Models using Predictive Modeling Markup Language

### 2014-11-30 paper.tex
Molecular descriptor representation

>>>>

\subsection{Representation of descriptors}

A molecular descriptor is a characteristic that is obtained by applying
a particular algorithm to a particular chemical structure representation.
Typically classified by the complexity of structure representation:
\begin{enumerate}
  \item 1D. Atom counts.
  \item 2D. Atom connectivity.
  \item 3D. Atom locations.
\end{enumerate}

Making descriptor calculation reproducible is difficult because it is
very tightly coupled with the preparation of chemical compounds to a suitable
chemical structure representation. For example, consider a simplest 2D
molecular descriptor that simply counts the number of aromcatic carbon
atoms. The functional definition of aromaticity is still a matter of 
ongoing debate -- classical organic chemistry interpretation (Hückel rule) 
vs. modern theoretical/computational interpretation.

By definition, when the algorithm is applied to an incorrect chemical
structure representation, the results is an incorrect molecular descriptor
value. Cannot reliably say if an user-specified representation is correct
or not by merely "looking" at it. Will need to apply a rigorous verification
procedure. For example, suppose SMILES chemical structure representation.
There exist different variations of SMILES, some of which make aromaticy
transparent by using different atom and bond symbols (ie. lowercase letter "c" 
and/or carbons connected by ":" bonds). Therefore, simple analysis of SMILES
using simple text analysis functions won't suffice, and it has to be parsed 
and reconstructed into more sophisticated chemical graph. Even then it is
possible to run into trouble, because the ring detection and ring aromaticity
assignment can be performed following different algorithms.

When "interfacing" QSAR models with actual end-users then it is necessary
to specify supported input chemical structure representations. Most likely,
actual chemical databases contain chemical compounds in data formats that
cannot be fed \emph{directly} to the molecular descriptor calculation
algorithm. For example, when testing with new chemical compounds then user
input may originate from graphical drawing board (eg. JMol applet [ref?] or 
J(S)ME Java(Script) Molecular Editor applet [ref?]) or be a systematic
name. The desirable property of a QSAR predictive solution would be to
recognize and automatically convert between different chemical structure
representations.

Unlike model training algorithm, the molecular descriptor calculation
algorithm does not have intermediary reusable states. From the practical
point of view, can be only reused by providing the complete algorithm in 
the application code form.

There have been attempts to "simplify" the systematizing and communicating 
algorithms using the ontology approach. The earliest work is the Blue Obelisk
Descriptor Ontology [ref?], which has been now superseded by the Chemical 
Information (CHEMINF) Ontology [ref?]. Unfortunately, not much adoption, 
because the vocabulary of real-life algorithms is way beyond the standardized
vocabularies, and the lack of competing implementations. An ontology that
has been crafted following CDK data structures and concepts is unlikely to
be relevant and useful for other descriptor calculation software. In other
words, if there is an one-to-one correspondence between ontology identifiers
and CDK implementation classes, then it may be a better idea to give those
CDK class names directly, and avoid the intermediary layers of "ontology 
identifiers" altogether.

The futility of formalizing algorithms using the ontology approach is
well demonstrated by the "parametrization problem". The behaviour of most
algorithms can be tweaked by adjusting a set of control variables. For 
example, when performing the analysis of 3D molecular shapes, finding the 
optimum between accuracy and speed (eg. mesh size). A different
parametrization will likely result in different calculated molecular
descriptor values. Cannot have repeatable nor less reproducible science
when the exact parametrization is unavailable. The only way to address
this problem would be to extend the algorithm ontology with this extra
information, but that would limit its relevancy and usefulness even more.

Another source of "variance" for molecular descriptor algorithms are 
programming errors. It is possible that newer versions of the same software 
could yield different results than older versions, because the algorithm
has been slightly changed (or reparameterized) in some point in time.
